---
title: "Your A4 Report Title (TODO: Revise title)"
author: "Lilli Toone"
date: '2022-11-01'
output: html_document
---

```{r echo=FALSE, eval=TRUE, message=FALSE}
source("../source/a4-helpers.R")
source("../source/analysis.R")
```

## Abstract

The proportions of different identities in jail and prison populations have long differed from community populations. This report visualizes data from the _Vera Institute_, an institution that has examined disparities in mass incarcerations. This report features the results of policy changes on the national prison population over time, possible disparities between female and male jail populations, and possible inequitable policies by counties near the U.S.-Mexico border.

## Introduction

The social issues related to the dataset are the perceptions of a value and possible obscelence. Community members and law enforcement may share the value of safety, but their definitions and perception of safety may differ. We have seen evidence of this with the 2020 _Black Lives Matter_ movements. If this difference becomes too great and these two parties fail to agree on policies for incarceration, the prison and jail system may become obsolete.  

Direct stakeholders of the dataset include people who are being assessed by local, state, and federal authorities and are at risk of going to jail. Indirect stakeholders of the dataset include people who live in areas nearby jails and prisons. These people are affected by the safety of their community and prevalence of authorities also in the area. Thus, the central values to this data set are safety and equity of community members before and after entering jail or prison.  

This project will examine potential outcomes of inequality in U.S. prisons. Some of the questions explored are:
1. How has the U.S. prison population changed over time?
2. Are state prison populations changing similarly, or are some states significantly contributing to the national change?
3. What is the observational variance between the proportion of female and male jail populations?
4. What is the geographical variance of jails that hold populations for U.S. Immigration and Customs Enforcement?

## Data Summary 

```{r echo=FALSE}
summary <- summary()
f_jail_mean <- pull(summary, f_mean)
m_jail_mean <- pull(summary, m_mean)
val_verde_ice <- pull(summary, val_verde_ice)
terrell_ice <- pull(summary, terrell_ice)
```

One key anomaly can be found between the female and male jail populations. The average jail population of women was **`r f_jail_mean` people**, and the average jail population of men was **`r m_jail_mean` people**. This difference, after some analysis, was found to be statistically significant. These variables highlight possible inequalities between the rate at which women and men are put into jail.  

An additional key phenomenon is that the population in jails for ICE vary significantly from county to county. For example, over the past decade, Terrell county in Texas held **`r terrell_ice` people** in jail for Immigration and Customs Enforcement. Meanwhile, the county just next door, Val Verde county, held **`r val_verde_ice` people** in jail for the exact same reason. This descrepancy suggests policies or a power imbalance that affects the amount of control ICE has over populations.

## Growth of the U.S. Prison Population
[TODO: >= 50 words]
- dplyr and ggplot2
- replicate figure 1 on canvas (increase of jail populations 1970-2018)
- create 2 functions;
- get_year_jail_pop() returns df suitable for vis.
- plot_jail_pop_for_us() returns chart. calls get_year_jail_pop
- caption chart
- summarize key patterns from chart
- concise, clear question
- labels, legend, scales

```{r echo=FALSE}
plot_jail_pop_for_us()
```

## Growth of Prison Population by State
[TODO: >= 50 words]
- dplyr and ggplot2
- line chart 1970~2018
- show by one or more states
- chart caption
- create 2 functions;
- get_jail_pop_by_states(states) takes in vector of states. returns a df for vis.
- plot_jail_pop_by_states(states) takes in a vector of states. should call get_jail_pop_by_states. returns chart.
- states <- 3~10 states
- concise, clear question
- labels, legend, scales

```{r echo=FALSE}
```

## Observational Variance
[TODO: >= 50 words]
- chart
- chart caption
- 2 continuous variables
- >=2 functions, one wrangling, one plotting (plot call wran)
- clear question
- summarize key pattern
- labels, legend, scales

```{r echo=FALSE}
```

## Geographical Variance
- chart
- chart caption
- 1 variable on a map
- >=2 functions, one wrangling, one plotting (plot call wran)
- clear question
- summarize key pattern
- labels, legend, scales
- Use a map based coordinate system to set the aspect ratio of your map

```{r echo=FALSE}
```
